Lessons Learned about JWST Transiting Planet Time Series from Ground-based Studies of the NIRCam Detectors

### 2020-04-19 ms.tex
typed of derivation of covariance weighted optimal flux

\begin{deluxetable*}{c|cc|cc}[b!]
\tablecaption{Summary of Noise Sources in OTIS Dark Test}
\tabletypesize{\footnotesize}
%\tablecolumns{7}
%\tablenum{2}
\tablewidth{0pt}
\tablehead{
\colhead{Aperture} &
\multicolumn{2}{c}{WLP8} &
\multicolumn{2}{c}{Grism 0.17 $\mu$m bin}\\
\colhead{Noise Description} &
\colhead{$\sigma$} &
\colhead{$\sigma$} &
\colhead{$\sigma$} &
\colhead{$\sigma$} \\
\colhead{} &
\colhead{DN} &
\colhead{ppm} &
\colhead{DN} &
\colhead{ppm}\\}
\startdata
Ideal Photon Noise & 4,400 & 113 & 1,420 & 390 \\
Ideal Uncorrelated Read Noise & 1,300 & 33 & 348 & 95 \\
\hline
Measured Read Noise with amplifier offsets (no correction) & 23,500 & 600 & 77,600 & 21,200  \\
Measured Read Noise (reference pixel correction) & 10,800 & 280 & 9,640 & 2,640 \\
Measured Read Noise (reference channel subtraction) & 9,700 & 250 & & \\
Measured Read Noise (row-by-row subtraction) & 6,000 & 150 & 3,630 & 992 \\
Measured Read Noise (20 Component PCA subtraction) & 5,400 & 140 & & \\
Measured Read Noise (smoothed 161 px kernel interpolating over aperture) & 3,300 & 85 & & \\
Measured Read Noise (5 Component PCA on each amplifier) & 3,200 & 81 & & \\
Measured Read Noise (row-by-row amp-by-amp mean subtraction) & 2,900 & 75 & & \\
Measured Read Noise (smoothed 40 px kernel on top of aperture) & 900 & 23 & & \\
Measured Read Noise (profile-weighted summation)	& & & 1,310 & 355 \\
\enddata
\tablecomments{The theoretical and measured noise sources in one integration for a Weak Lens +8 and Grism Time Series.
The ppm units are assuming a peak 60\% well depth or 3.9$\times 10^7$ DN during an integration for the weak lens PSF and 3.7$\times 10^6$ DN during and integration for the grism PSF.
The noise is reported for the the aperture/spectral extraction (in DN) or relative to the photon signal (in ppm).
The integration is assumed to be a bright source with 2 groups of 1 integration each up the ramp (Correlated Double Sampling or CDS mode).}
\label{tab:noiseSummary1overf}
\end{deluxetable*}

@@

We give some illustrative numbers for a 0.165 $\mu$m wide bin of a grism spectrum centered on 3.25 $\mu$m as listed in Table \ref{tab:noiseSummary1overf}.
Without reference pixel subtraction, the grism apertures have standard deviations of 77,600 DN.
Reference pixel subtraction drops this value to 9,640 DN and row-by-row subtraction drops it to 3,630 DN.
For a well depth of 60\%, there are $3.7 \times 10^6$ DN of source counts in the 0.17 $\mu$m bin, which amounts to a photon error of 1,420 DN assuming a gain of 1.84.
Therefore, 1/f noise can potentially dominate the error budget for broadband grism time series!
Expressed in ppm, the read noise after processing with row-by-row subtraction is 990 ppm as compared to a photon noise of 95 ppm for a CDS subtraction (2 groups in ramp).

@@

Given the high level of correlation between pixels, another way to reduce 1/f noise is with a small aperture.
For example, a 15 pixel wide aperture results in measured read noise of 990 ppm for the same 0.17$\mu$m wide bin centered at 4.25$\mu$m whereas a 3 pixel wide aperture results in a measured read noise of 290 ppm, which is below the expected photon noise.
For a one pixel wide aperture, the measured read noise is only 230 ppm.
Of course such small apertures will be subject to other losses by missing so much of the PSF.
Therefore, the spectroscopic pipeline should either use a small aperture or provide extra weights to the brightest pixels such as in optimal extraction \citep[e.g.][]{horne1986optimalE}.
If one calculates the weights to the spectra simply by the signal-to-noise while assuming each pixel's noise is independent, the weights are not heavy enough for the brightest pixels.
This is because the observations will largely be in the foreground limit (where the variance is equal to the number of photons), where weighting by the profile for pixel $j$ (P$_j$) divided by the variance (V$_i$) becomes
\begin{equation}
w_j = P_j / V_j = N_{ph} / N_{ph} = 1.0,
\end{equation}
where $N_{ph}$ is the number of photons on that pixel.

However, if one generalizes the weighted average to a covariance matrix \citep{schmelling1995averagingCorrelatedData}, the weights become,
\begin{equation}
w_j = P_j \sum_{i=1}^{m} \left( C^{-1}\right)_{ij},\label{eq:covarianceWeighted}
\end{equation}
where $C^{-1}$ is the inverse of the covariance matrix between pixels $i$ and $j$ and $m$ is the number of pixels in the summation.
The optimally weighted flux ($f_{opt}$) then becomes
\begin{equation}
f_{opt} = \frac{\sum_{j=1}^{m} w_j S_j}{\sum_{j=1}^{m} w_j P_j},
\end{equation}
where $S_j$ is the background subtracted flux.

% the longer form I'm not using
%the weighted flux becomes,
%\begin{equation}\label{eq:covarianceWeighted}
%f_{opt} = \frac{\sum_{i,j=0}^{m} C^{-1}_{ij} S_j P_j}{\sum_{i,j=0}^{m} C^{-1}_{ij} P_j^2},
%\end{equation}
%where $C^{-1}$ is the inverse of the covariance matrix between pixels $i$ and $j$ and $m$ is the number of pixels in the summation and $S_j$ is the background subtracted flux.

Simply variance weighting the spectra by the read and photon noise does not provide heavy enough weights.
If we assume the read noise is perfectly correlated along the spectral direction and uncorrelated along the spatial direction, then equation \ref{eq:covarianceWeighted} becomes
\begin{equation}
w_j = \frac{P_j}{N_{ph} + N_{px} \sigma_R^2},
\end{equation}
where $N_{px}$ is the number of spectral pixels that will be binned and $\sigma_R$ is the read noise per pixel.
Thus, for large values of $N_{px}$ (up to 512 for an amplifier), the $\sigma_R \approx 14e^-$ of read noise per pixel, can compound to 100,400 e$^-$ and the weighting approaches
\begin{equation}
w_j \propto P_j.
\end{equation}

Profile weighting should therefore lower the error on wavelength-binned time series.
For the grism aperture experiment, we find that profile weighting gives a measured read noise of 355 ppm.
Thus, profile weighting still does not reduce the standard deviation over an extraction aperture that is just 1 pixel wide at the peak grism flux.
We show the real covariance matrix in Section \ref{sec:empCovarianceMatrix}.

\subsubsection{Empirical Covariance Matrix}\label{sec:empCovarianceMatrix}
The ideal weighting of the spectrum from Equation \ref{eq:covarianceWeighted} depends on the covariance matrix $C_{ij}$.
We use CDS (correlated double sampling) read pairs of OTIS long dark data (108 frames) with reference pixel corrections to estimate $C_{ij}$.
We consider a wavelength bin that is 100 pixels in the dispersion direction (0.1$\mu$m) and 14 spatial spatial pixels.
Figure \ref{fig:covSpecSpatial} shows the covariance matrix of this wavelength bin, taken from the median of 20 regions of the detector.
The diagonal elements are consistent with published NIRCam detector performance tables with $\sigma^2 \approx 13^2 (e^-)^2=169 (e^-)^2$.
It is clear that 1/f noise causes covariance most prominently along the spectral direction (the fast-read direction), with a covariance nearly uniformly at 36~$(e^-)^2$.
Surprisingly, there is also a covariance between spatial pixels (the slow read direction) of $\sim$16 $(e^-)^2$ that falls off to $\sim 10 (e^-)^2$.

>>>>

\begin{deluxetable*}{c|cc|cc}[b!]
\tablecaption{Summary of Noise Sources in OTIS Dark Test}
\tabletypesize{\footnotesize}
%\tablecolumns{7}
%\tablenum{2}
\tablewidth{0pt}
\tablehead{
\colhead{Aperture} &
\multicolumn{2}{c}{WLP8} &
\multicolumn{2}{c}{Grism 0.17 $\mu$m bin}\\
\colhead{Noise Description} &
\colhead{$\sigma$} &
\colhead{$\sigma$} &
\colhead{$\sigma$} &
\colhead{$\sigma$} \\
\colhead{} &
\colhead{DN} &
\colhead{ppm} &
\colhead{DN} &
\colhead{ppm}\\}
\startdata
Ideal Photon Noise & 4,400 & 113 & 1,420 & 390 \\
Ideal Uncorrelated Read Noise & 1,300 & 33 & 348 & 95 \\
\hline
Measured Read Noise with amplifier offsets (no correction) & 23,500 & 600 & 77,600 & 21,200  \\
Measured Read Noise (reference pixel correction) & 10,800 & 280 & 9,640 & 2,640 \\
Measured Read Noise (reference channel subtraction) & 9,700 & 250 & & \\
Measured Read Noise (row-by-row subtraction) & 6,000 & 150 & 3,630 & 992 \\
Measured Read Noise (20 Component PCA subtraction) & 5,400 & 140 & & \\
Measured Read Noise (smoothed 161 px kernel interpolating over aperture) & 3,300 & 85 & & \\
Measured Read Noise (5 Component PCA on each amplifier) & 3,200 & 81 & & \\
Measured Read Noise (row-by-row amp-by-amp mean subtraction) & 2,900 & 75 & & \\
Measured Read Noise (smoothed 40 px kernel on top of aperture) & 900 & 23 & & \\
Measured Read Noise (profile-weighted summation)	& & & 1,310 & 355 \\
\enddata
\tablecomments{The theoretical and measured noise sources in one integration for a Weak Lens +8 and Grism Time Series.
The ppm units are assuming a peak 60\% well depth or 3.9$\times 10^7$ DN during an integration for the weak lens PSF and 3.7$\times 10^6$ DN during and integration for the grism PSF.
The noise is reported for the the aperture/spectral extraction (in DN) or relative to the photon signal (in ppm).
The integration is assumed to be a bright source with 2 groups of 1 integration each up the ramp (Correlated Double Sampling or CDS mode).
For the grism, statistics are shown for a representative 0.17~$\mu$m wide wavelength bin centered on 3.34~$\mu$m.}
\label{tab:noiseSummary1overf}
\end{deluxetable*}

@@

We give some illustrative numbers for a 0.165 $\mu$m wide bin of a grism spectrum centered on 3.34 $\mu$m as listed in Table \ref{tab:noiseSummary1overf}.
Without reference pixel subtraction, the grism apertures have standard deviations of 77,600 DN.
Reference pixel subtraction drops this value to 9,640 DN and row-by-row subtraction drops it to 3,630 DN.
For a well depth of 60\%, there are $3.7 \times 10^6$ DN of source counts in the 0.17 $\mu$m bin, which amounts to a photon error of 1,420 DN assuming a gain of 1.84.
Therefore, 1/f noise can potentially dominate the error budget for broadband grism time series!
Expressed in ppm, the read noise after processing with row-by-row subtraction is 990 ppm as compared to a photon noise of 95 ppm for a CDS subtraction (2 groups in ramp).

@@

Given the high level of correlation between pixels, another way to reduce 1/f noise is with a small aperture.
For example, a 15 pixel wide aperture results in measured read noise of 990 ppm for the same 0.17$\mu$m wide bin centered at 3.34$\mu$m whereas a 3 pixel wide aperture results in a measured read noise of 290 ppm, which is below the expected photon noise.
For a one pixel wide aperture, the measured read noise is only 230 ppm.
Of course such small apertures will be subject to other losses by missing so much of the PSF and the photon noise will be a larger fraction of the flux.
Therefore, a weighting scheme expanded from optimal extraction \citep{horne1986optimalE} is preferred.

\subsubsection{Optimally Co-Variance Weighted Flux}\label{sec:optimalCovWeights}

Here, we provide a modification to the optimal extraction formula from \citet{horne1986optimalE}, that uses the covariance matrix of correlated pixels.
The goal is to find a scheme that estimates the total flux by summing the value in each pixel $j$.
We start by calculating a profile function $P_j$ that describes the fraction of the total flux in pixel $j$.
The profile is normalized so that
\begin{equation}
\sum_{j=1}^{m} P_j = 1.
\end{equation}
Any pixel is can be used to estimate the total flux $f_i$ by dividing by the profile.
\begin{equation}
f_j = \frac{S_j}{P_j},
\end{equation}
where $S_j$ is the background-subtracted flux.
Therefore, the flux total is a weighted average of each pixel's estimate of the total flux $f_j$.
The goal becomes a weighting scheme to determine which weights $w_j$ minimize the variance in the total background estimate:
\begin{equation}
f_{opt} = \frac{\sum_{j=1}^{m} w_j \frac{S_j}{P_j}}{\sum_{j=1}^{m} w_j}.
\end{equation}

If one assumes that all pixels are independent (and have zero correlation), the optimal weights (ones that minimize the variance in $f_{opt}$) are
\begin{equation}
w_j = \frac{1}{\mathrm{Var}(\frac{S_j}{P_j})} = \frac{1}{\frac{1}{P_j^2} \mathrm{Var}(S_j)} = \frac{P_j^2}{V_j},
\end{equation}
where Var() is the variance or the expectation of the deviation from the mean and $V_j =$Var$(S_j$) \citep{horne1986optimalE}.
Exoplanet observations, where high signal to noise is needed to remove the planet from star signal, are generally in the foreground limit (where the variance is approximately equal to the number of photons).
In this case, both the profile $P_j$ and variance $V_j$ will be proportional to the number of photons $N_{phot}$, so the pre-multiplier of the flux become
\begin{equation}
\frac{w_j}{P_j} \approx \frac{N_{phot}^2}{N_{phot} N_{phot}} = 1.0.
\end{equation}
Therefore, using this traditional formula will provide little difference with straight summation,
\begin{equation}
f_{sum} = \sum_{j=1}^{m} S_j.
\end{equation}

However, in the presence of noise correlations, the formulae become modified.
The optimal weights for correlated data are \citep{schmelling1995averagingCorrelatedData} the sum of the inverse covariance matrix,
\begin{equation}\label{eq:covarianceweights}
w_j = \sum_{i=0}^{m} C_{ij}^{-1},
\end{equation}
where $C_{ij}$ is the covariance (Cov()) of the variable $S_i/P_i$ with $S_j/P_j$,
\begin{equation}
C_{ij} = \mathrm{Cov}\left(\frac{S_i}{P_i},\frac{S_j}{P_j}\right) = \frac{1}{P_i P_j} \mathrm{Cov}(S_i,S_j).
\end{equation}

%Simply variance weighting the spectra by the read and photon noise does not provide heavy enough weights.
%If we assume the read noise is perfectly correlated along the spectral direction and uncorrelated along the spatial direction, then equation \ref{eq:covarianceweights} becomes
%\begin{equation}
%w_j = \frac{P_j}{N_{ph} + N_{px} \sigma_R^2},
%\end{equation}
%where $N_{px}$ is the number of spectral pixels that will be binned and $\sigma_R$ is the read noise per pixel.
%Thus, for large values of $N_{px}$ (up to 512 for an amplifier), the $\sigma_R \approx 14e^-$ of read noise per pixel, can compound to 100,400 e$^-$ and the weighting approaches
%\begin{equation}
%w_j \propto P_j.
%\end{equation}
%
%Profile weighting should therefore lower the error on wavelength-binned time series.
%For the grism aperture experiment, we find that profile weighting gives a measured read noise of 355 ppm.
%Thus, profile weighting still does not reduce the standard deviation over an extraction aperture that is just 1 pixel wide at the peak grism flux.
%We show the real covariance matrix in Section \ref{sec:empCovarianceMatrix}.

\subsubsection{Empirical Covariance Matrix}\label{sec:empCovarianceMatrix}
The ideal weighting of the spectrum from Equation \ref{eq:covarianceweights} depends on the covariance matrix $C_{ij}$.
We use CDS (correlated double sampling) read pairs of OTIS long dark data (108 frames) with reference pixel corrections to estimate $C_{ij}$.
We consider a wavelength bin that is 100 pixels in the dispersion direction (0.1$\mu$m) and 14 spatial spatial pixels.
Figure \ref{fig:covSpecSpatial} shows the covariance matrix of this wavelength bin, taken from the median of 20 regions of the detector.
The diagonal elements are consistent with published NIRCam detector performance tables with $\sigma^2 \approx 13^2 (e^-)^2=169 (e^-)^2$.
It is clear that 1/f noise causes covariance most prominently along the spectral direction (the fast-read direction), with a covariance nearly uniformly at 36~$(e^-)^2$.
Surprisingly, there is also a covariance between spatial pixels (the slow read direction) of $\sim$16 $(e^-)^2$ that falls off to $\sim 10 (e^-)^2$.
